\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-08-06 writing/manuscript.Rmd
addressed Aimee's comments (1); corrected typos

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].

@@

\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** an animal can access. Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, *de-novo* findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs. In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little to no information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use. For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies and life histories being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differents in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

@@

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.

There are many distributions we can assign to $R$, depending on how we quantify it. For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, since there is no commonly used distribution with that support, the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).

@@

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

@@

Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time. Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$. Secondly, the animal's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

@@

The designation of high-quality habitats is a central concern of many conservation initiatives. This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat. With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas. In light of the UN's objctive to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].

@@

It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels. While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity. In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants. Consequently, the abundance of primary consumers may correlate with productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019]. Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay. Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity. However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less sotchastic resources.

@@

>>>>

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].

@@

\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** the animal can access. Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs. In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and difficult to quantify), since it depends on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

@@

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Bernoulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.

There are many distributions we can assign to $R$ depending on how we quantify it. For instance, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework). Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.

@@

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

@@

Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time. Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$. Secondly, the animal's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

@@

The designation of high-quality habitats is a central concern of many conservation initiatives. This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat. With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas. In light of the UN's objective to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].

@@

It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels. While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity. In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants. Consequently, the abundance of primary consumers may correlate with productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019]. Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay. Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity. However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less stochastic resources.

@@

# Acknowledgements

We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project. In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
